### 2015-06-05 chapters/encoding.tex
wraps line at character width 80 in all tex files

There are many in-depth histories of the origin and development of writing systems (e.g.~\citet{Robinson1997,Powell2012}), a story that we will not repeat here. However, the history of turning writing into computer readable code is not so often told, so we decided to offer a short survey of the major developments here. This history turns out to be intimately related to the history of telegraphic communication.\footnote{Because of the recent history as summarized in this section, we have used mostly rather ephemeral internet sources. When not references by traditional literature in the bibliography, we have specifically used \url{http://www.unicode.org/history/} and various Wikipedia pages for the information presented here. A useful survey of the historical development of the physical hardware of telegraphy and telecommunication is \citet{Huurdeman2003}. Most books that discuss the development of encoding of telegraphic communication focus of cryptography, e.g.~\citet{Singh1999}, and forego the rather interesting story of `open' encoding that is related here.}

@@

In the context of electric telegraphy, such a binary code system was first established by Émile Baudot in 1870, using a fixed combination of five signals for each written character.\footnote{True binary codes have a longer history, going at least back to the Baconian cipher devised by Francis Bacon in 1605. However, the proposal by Baudot was the quintessential proposal leading to all modern systems.} There are $2^5 = 32$ possible combination when using five binary signals; an encoding today designated as \textsc{5-bit}. These codes are sufficient for all Latin letters, but of course they do not suffice for all written symbols, including punctuation and digits. As a solution, the Baudot code uses a so-called \textsc{shift} character, which signifies that from that point onwards---until shifted back---a different encoding is used, allowing for yet another set of 32 codes. In effect, this means that the Baudot code, and the \textsc{International Telegraph Alphabet} (ITA) derived from it, had an extra `bit' of information, so the encoding is actually 6-bit (with $2^6 = 64$ different possible characters). For decades, this encoding was the standard for all telegraphy and it is still in limited use today.

>>>>

There are many in-depth histories of the origin and development of writing
systems (e.g.~\citet{Robinson1997,Powell2012}), a story that we will not repeat
here. However, the history of turning writing into computer readable code is not
so often told, so we decided to offer a short survey of the major developments
here. This history turns out to be intimately related to the history of
telegraphic communication.\footnote{Because of the recent history as summarized
in this section, we have used mostly rather ephemeral internet sources. When not
references by traditional literature in the bibliography, we have specifically
used \url{http://www.unicode.org/history/} and various Wikipedia pages for the
information presented here. A useful survey of the historical development of the
physical hardware of telegraphy and telecommunication is \citet{Huurdeman2003}.
Most books that discuss the development of encoding of telegraphic communication
focus of cryptography, e.g.~\citet{Singh1999}, and forego the rather interesting
story of ``open encoding'' that is related here.}

@@

In the context of electric telegraphy, such a binary code system was first
established by Émile Baudot in 1870, using a fixed combination of five signals
for each written character.\footnote{True binary codes have a longer history,
going at least back to the Baconian cipher devised by Francis Bacon in 1605.
However, the proposal by Baudot was the quintessential proposal leading to all
modern systems.} There are $2^5 = 32$ possible combination when using five
binary signals; an encoding today designated as \textsc{5-bit}. These codes are
sufficient for all Latin letters, but of course they do not suffice for all
written symbols, including punctuation and digits. As a solution, the Baudot
code uses a so-called \textsc{shift} character, which signifies that from that
point onwards---until shifted back---a different encoding is used, allowing for
yet another set of 32 codes. In effect, this means that the Baudot code, and the
\textsc{International Telegraph Alphabet} (ITA) derived from it, had an extra
\textsc{bit} of information, so the encoding is actually 6-bit (with $2^6 = 64$
different possible characters). For decades, this encoding was the standard for
all telegraphy and it is still in limited use today.

### 2015-06-05 chapters/introduction.tex
wraps line at character width 80 in all tex files

When considering the worldwide linguistic diversity, including all lesser-studied and endangered languages, there exist numerous different orthographies using symbols from the same scripts. For example, there are hundreds of orthographies using Latin-based alphabetic scripts. All of these orthographies use the same symbols, but these symbols differ in meaning and usage throughout the various orthographies. To be able to computationally use and compare different orthographies, we need a way to specify all orthographic idiosyncrasies in a computer-readable format (a process called `tailoring' in Unicode parlance). We call such specifications \textsc{orthography profiles}. Ideally, these specifications have to be integrated into so-called Unicode locale descriptions, though we will argue that in practice this is often not the most useful solution for the kind of problems arising in the daily practice of linguistics. Consequently, a central goal of this paper is to flesh out the linguistic challenges for locale descriptions, and work out suggestions to improve their structure for usage in a linguistic context. Conversely, we also aim to improve linguists' understanding and appreciation for the accomplishments of the Unicode Consortium in the development of the Unicode Standard.

The necessity to computationally use and compare different orthographies most forcefully arises in the context of language comparison. Concretely, in our current research our goal is to develop quantitative methods for language comparison and historical analysis in order to investigate worldwide linguistic variation and to model the historical and areal processes that underlie linguistic diversity, cf.~\citet{Steiner_etal2011,List2012,List2012a,ListMoran2013,MoranProkic2013}. In this work, it is crucial to be able to flexibly process across numerous resources with different orthographies. In many cases even different resources on the `same' language use different orthographic conventions. Another orthographic challenge that we encounter regularly in our linguistic practice is electronic resources on a particular language that claim to follow a specific orthographic convention (often a resource-specific convention), but on closer inspection such resources are almost always not consistently encoded. Thus a second goal of our orthography profiles is to allow for an easy specification of orthographic conventions, and use such profiles to check consistency and to report errors to be corrected.

>>>>

When considering the worldwide linguistic diversity, including all
lesser-studied and endangered languages, there exist numerous different
orthographies using symbols from the same scripts. For example, there are
hundreds of orthographies using Latin-based alphabetic scripts. All of these
orthographies use the same symbols, but these symbols differ in meaning and
usage throughout the various orthographies. To be able to computationally use
and compare different orthographies, we need a way to specify all orthographic
idiosyncrasies in a computer-readable format (a process called
\textsc{tailoring} in Unicode parlance). We call such specifications
\textsc{orthography profiles}. Ideally, these specifications have to be
integrated into so-called Unicode locale descriptions, though we will argue that
in practice this is often not the most useful solution for the kind of problems
arising in the daily practice of linguistics. Consequently, a central goal of
this paper is to flesh out the linguistic challenges for locale descriptions,
and work out suggestions to improve their structure for usage in a linguistic
context. Conversely, we also aim to improve linguists' understanding and
appreciation for the accomplishments of the Unicode Consortium in the
development of the Unicode Standard.

The necessity to computationally use and compare different orthographies most
forcefully arises in the context of language comparison. Concretely, in our
current research our goal is to develop quantitative methods for language
comparison and historical analysis in order to investigate worldwide linguistic
variation and to model the historical and areal processes that underlie
linguistic diversity,
cf.~\citet{Steiner_etal2011,List2012,List2012a,ListMoran2013,MoranProkic2013}.
In this work, it is crucial to be able to flexibly process across numerous
resources with different orthographies. In many cases even different resources
on the \textit{same} language use different orthographic conventions. Another
orthographic challenge that we encounter regularly in our linguistic practice is
electronic resources on a particular language that claim to follow a specific
orthographic convention (often a resource-specific convention), but on closer
inspection such resources are almost always not consistently encoded. Thus a
second goal of our orthography profiles is to allow for an easy specification of
orthographic conventions, and use such profiles to check consistency and to
report errors to be corrected.
